# Reproducible Research: Peer Assessment 1


## Loading and preprocessing the data

```{r}
if(!file.exists("activity.csv")) unzip("activity.zip")
activity <- read.csv("activity.csv", colClasses=c("integer", "Date", "integer"))
head(activity)
```

## What is mean total number of steps taken per day?

```{r}
totals <- with(activity, tapply(steps, date, sum, na.rm=F))
hist(totals, main="Total number of steps taken each day", xlab="Steps", ylab="Number of days", axes=F)
axis(1, seq(0, 25000, 5000), labels=prettyNum(seq(0, 25000, 5000), big.mark=",", scientific=F))
axis(2, seq(0, 25, 5))
m1 <- mean(totals, na.rm=T)
m2 <- median(totals, na.rm=T)
abline(v=m1, col="red")
```

The mean total number of steps taken per day is `r prettyNum(m1, big.mark=",", scientific=F)` (indicated by the red bar in the plot above) and the median is `r prettyNum(m2, big.mark=",", scientific=F)`.

## What is the average daily activity pattern?

```{r}
means <- with(activity, tapply(steps, interval, mean, na.rm=T))
plot(as.integer(names(means)), means, type="l", xlab="Interval", ylab="Steps", main="Average number of steps taken per 5-minute interval across all days")
means.max <- which.max(means)
```

The maximum average number of steps taken per 5-minute interval across all days was `r means.max` and occurred in interval `r names(means.max)`.

## Imputing missing values

We shall impute missing values using the average number of steps taken per 5-minute interval across all days.

```{r}
new.activity <- activity
for(i in 1:nrow(new.activity)) 
        new.activity$steps[i] <- ifelse(is.na(new.activity$steps[i]), means[as.character(new.activity$interval[i])], new.activity$steps[i])
head(new.activity)
new.totals <- with(new.activity, tapply(steps, date, sum, na.rm=F))
hist(new.totals, main="Total number of steps taken each day", xlab="Steps", ylab="Number of days", axes=F)
axis(1, seq(0, 25000, 5000), labels=prettyNum(seq(0, 25000, 5000), big.mark=",", scientific=F))
axis(2, seq(0, 25, 5))
new.m1 <- mean(new.totals, na.rm=T)
new.m2 <- median(new.totals, na.rm=T)
abline(v=new.m1, col="blue")
```

After imputting missing values in this way, the mean total number of steps taken per day is `r prettyNum(new.m1, big.mark=",", scientific=F)` (indicated by the blue bar in the plot above) and the median is `r prettyNum(new.m2, big.mark=",", scientific=F)`, i.e., only the median is a little changed.

## Are there differences in activity patterns between weekdays and weekends?

```{r}
new.activity <- transform(new.activity, weekday = weekdays(date, abbreviate=T))
new.activity <- transform(new.activity, weekday = sub("Sat|Sun", "weekend", weekday))
new.activity <- transform(new.activity, weekday = sub("Mon|Tue|Wed|Thu|Fri", "weekday", weekday))
new.activity$weekday <- as.factor(new.activity$weekday)
new.means <- with(new.activity, tapply(steps, list(weekday, interval), mean))
par(mfrow=c(2,1))
plot(names(new.means[2,]), new.means[2,], type="l", xlab="Interval", ylab="Steps", main="Average number of steps taken per interval during weekends", ylim = c(0, 250))
plot(names(new.means[1,]), new.means[1,], type="l", xlab="Interval", ylab="Steps", main="Average number of steps taken per interval during weekdays", ylim = c(0, 250))
```

The average number of steps taken per interval during the earlier part of the day is lower on weekends than on weekdays. This is perhaps due to the absence of a commute during weekends. The average numnber of steps taken per interval during the later of the day is higher on weekends than on weekdays. This is perhaps due to not being tied to a desk during weekends.  The average total number of steps taken is `r prettyNum(sum(new.means[2,]), big.mark=",", scientific=F)` during weekends and `r prettyNum(sum(new.means[1,]), big.mark=",", scientific=F)` during weekdays.